---
title: "class12"
author: "Marielle Samonte (A16861951)"
format: pdf
---

# Section 1. Proportion og G/G in a population

Download a CSV file from Emsemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39889641-39898742;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Lets now dig into this further.

# Section 4. Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
```{r}
summary(expr)
```

Sample size for each genotype:
```{r}
table(expr$geno)
```
Median expression levels for each genotype:
```{r}
round( tapply(expr$exp, expr$geno, median), 2 )
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr) + 
  aes(geno, exp, fill = geno) +
  geom_boxplot()
```
G/G has a lower median and overall lower expression of the ORMDL3 gene than A/A. This indicates that the SNP effects the expression of ORMDL3, specifically the G allele in this location is associated with reduced expression of the ORMDL3 gene.

