---
title: "Class 5: Data Viz with ggplot"
author: "Marielle Samonte (PID: A16861951)"
format: pdf
toc: true
---

# Background

There are many graphics systems available in R. These include "base" R and tones of add on packages like **ggplot2**.

Let's compare "base" and **ggplot2** briefly. We can use some example data that is built-in with R called `cars`:

```{r}
head(cars)
```

In base R I can just call `plot()`

```{r}
plot(cars)
```

How can we do this with **ggplot2**

First we need to install the package. We do this `install.packages("ggplot2")`. I only need to do this once and then it will be available on my computer from then on. 

> Key point: I only install packages in the R console not within quarto docs or R scripts. 

Before I use any add-on package I must load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot has at least 3 things:

- the **data** (in our case `cars`)
- the **aes**thetics (how the data map to the plot)
- the **geom**s that determine how the plot is drawn (lines, points, columns, etc.)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist)
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

For "simple" plots ggplot is much more verbose than base R but the defaults are nicer and for complicated plots it becomes much more efficient and structured.

> Q. Add a line to show the relationship of speed to stopping distance (i.e. add another "layer")

```{r}
p <- ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(se=FALSE, method="lm")

```

I can always save any ggplot object (i.e. plot) and then use it later for adding more layers.

```{r}
p
```

> Q. Add a title and subtitle to the plot 

```{r}
p + labs(title="My first ggplot",
         subtitle="Stopping distance of old cars", 
         caption="BIMM143", 
         x="Speed (MPG)", 
         y="Stopping distance (ft)") +
  theme_bw()

```

# Gene expression plot

Read input data into R

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q. How many genes are in ths wee dataset?

```{r}
nrow(genes)
```

> Q. How many columns are there?

```{r}
ncol(genes)
```

> Q. What are the column names?

```{r}
colnames(genes)
```

> Q. How many "up" and "down" regulated genes are there?

```{r}
table(genes$State)
```


## Custom color plot

> Q. Make a first plot of this data

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  scale_color_manual( values=c("red", "grey", "blue")) +
  geom_point() + 
  labs(title="Gene expression changes upon drug treatment", 
       x="Control (no drug)",
       y="Drug treated") +
  theme_bw()
```


# Using different geoms

Let's plot some aspects of the in-built `mtcars` dataset.

```{r}
head(mtcars)
```

> Q. Scatter plot of `mpg` vs `disp`

```{r}
p1 <- ggplot(mtcars) +
  aes(x=mpg, y=disp) +
  geom_point()

p1
```


> Q. boxplot of `gear` vs `disp`

```{r}
p2 <- ggplot(mtcars) +
  aes(x=gear, y=disp, group=gear) +
  geom_boxplot()

p2
```

> Q. barplot of `carb` 

```{r}
p3 <- ggplot(mtcars) +
  aes(carb) +
  geom_bar()

p3
```

> Q. Smooth of `disp` vs `qsec`

```{r}
p4 <- ggplot(mtcars) +
  aes(disp, qsec) +
  geom_smooth()

p4
```

I want to combine all these plots into one figure with multiple pannels. 

We can use the **patchwork** package to do this.

```{r}
library(patchwork)

(p1 / p2 | p3 / p4)
```

```{r}
ggsave(filename="myplot.png", width=5, height=3)
```


## Faceting

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

And a wee peak
```{r}
head(gapminder)
```

< Q. How many countries are in this dataset?

```{r}
length(table(gapminder$country))
```

> Q. Plot gdpPercap vs lifeExp color by continent

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(aplha=0.3) +
  facet_wrap(~continent) +
  theme_bw()
```



